---
title: "SPATA2_Object_initialisation_Tests"
output:
  html_document:
    df_print: paged
    output_dir: 
    output_file: 
  pdf_document:
    extra_dependencies: subfig
    latex_engine: xelatex
header-includes:
- \usepackage{fontspec}
- \setmainfont{Times New Roman}
---

## Setup + Loading Required Libraries:

Ensure Seurat V = 5.0.3 & SPATA2 V = 2.0.4

```{r setup, include=FALSE}
library(Seurat)
library(SeuratData)
library(ggplot2)
library(patchwork)
library(tidyverse)
library(dplyr)
library(spacexr)
library(SPATA2)
library(stringr) # for imaging
library(BayesSpace)
library(keras)
library(readr)
library(glmGamPoi)
library(gridExtra)
library(pheatmap)
library(RColorBrewer)

# Remember to change the output_file parameter in the YAML front matter
# according to the dataset used


## So that sufficient time is available to load the data from SPATAData
getOption("timeout")
options(timeout = 100)

# To avoid Warning message during knit to html:
# ggrepel: 8 unlabeled data points (too many overlaps). 
# Consider increasing max.overlaps
options(ggrepel.max.overlaps = Inf)


```

## There can be many different ways we can initiate the SPATA2 object:

Note: Regardless of which way object is initiated, the stress and mitochondrial genes are removed; if you want control over that probably need to create Seurat Object first and then do it.

-   Plain: Just the count matrix
-   LogNormalised: Count matrix gets log normalised,
-   SCTransform: Removes mitochondrial genes; Returns a matrix of pearson residuals, not log normalised; Turn off log normalisation steps when using this. 

```{r setting dataset for spata2 object}


# This is how a 10x Dataset from https://www.10xgenomics.com/datasets
# can be used as a SPATA2 object

# dataset_10x_dir has to be changed depending on where the dataset is stored
# Make sure you are in the right directory with the required files
# Ensure that the .h5 is named as "filtered_feature_bc_matrix.h5"

master_script_rmd_dir <- getwd()
dataset_10x_dir <- file.path(master_script_rmd_dir,"10x_Datasets/MMS_Datasets/GSE188888_Heart_2_samples/GSM5691529/Sample_G21d_181A8_2/outs")

# Alter the sample name value as required
# Preferably named well so that it can be easily identified

sample_name_spata2_obj <- "GSM5691529_Sample_G21d"


```


## plain spata2 obj; no normalisation

```{r plain spata2 obj}
# Alter the name of the obj for easy identification

spata2_obj_10x_2_plain <-
    initiateSpataObject_10X(
        directory_10X = dataset_10x_dir, # the directory from which to load the data
        sample_name = sample_name_spata2_obj,
        SCTransform = FALSE, # can be FALSE or diff value
        NormalizeData = FALSE,
        FindVariableFeatures = FALSE,
        ScaleData = FALSE
    )


# Assuming your sparse matrix object is called 'counts_matrix'
library(Matrix)

# Extract the non-zero counts
non_zero_counts <- spata2_obj_10x_2_plain@data[["GSM5691529_Sample_G21d"]][["counts"]]@x

# Plot the distribution of the non-zero counts
hist(non_zero_counts, breaks = 10000, main = "Distribution of Non-Zero Counts", xlab = "Counts", ylab = "Frequency", xlim = c(0, 30), ylim = c(0,1000000))

# If you want to see the overall distribution including zeros (not recommended for very sparse matrices due to high memory usage):
## There wont be any zeros as they were removed when the object was created
dense_counts_matrix <- as.matrix(spata2_obj_10x_2_plain@data[["GSM5691529_Sample_G21d"]][["counts"]])

#head(dense_counts_matrix)
all_counts <- rowMeans2(dense_counts_matrix)
hist(all_counts, breaks = 20000, main = "Distribution of All genes Counts Means", xlab = "Counts", ylab = "Frequency", xlim = c(0, 3))
count_df <- as.data.frame(dense_counts_matrix)

all_counts_vector <- as.vector(all_counts)
all_counts_df <- data.frame(Gene = names(all_counts), Counts = as.numeric(all_counts))
all_rows_sums_vector <- rowSums(count_df)
hist(all_rows_sums_vector, breaks = 10000000, main = "Distribution of All genes count Sums", xlab = "Counts", ylab = "Frequency", xlim = c(0, 10))

```
## just log normalised

```{r log normalised}

spata2_obj_10x_2_log_normalised <-
    initiateSpataObject_10X(
        directory_10X = dataset_10x_dir, # the directory from which to load the data
        sample_name = sample_name_spata2_obj,
        SCTransform = FALSE, # can be FALSE or diff value
        NormalizeData = TRUE,
        FindVariableFeatures = TRUE,
        ScaleData = TRUE
    )

hist(as.vector(spata2_obj_10x_2_log_normalised@data[["GSM5691529_Sample_G21d"]][["scaled"]]), breaks = 100, main = "Distribution of log Normalized Counts")

exp_matrix_log_normalised <- getExpressionMatrix(spata2_obj_10x_2_log_normalised)

# head(exp_matrix_log_normalised)
```

## log normalised and Sctransform is TRUE

```{r log normalised and Sctransform is TRUE}

# Does sct affect the log distribution
# This is what i used for my analysis

spata2_obj_10x_2_log_normalised_sct <-
    initiateSpataObject_10X(
        directory_10X = dataset_10x_dir, # the directory from which to load the data
        sample_name = sample_name_spata2_obj,
        SCTransform = TRUE, # can be FALSE or diff value
        NormalizeData = TRUE,
        FindVariableFeatures = TRUE,
        ScaleData = TRUE
    )

spata2_obj_10x_2_log_normalised_sct <- computeGeneMetaData(spata2_obj_10x_2_log_normalised_sct) 


hist(as.vector(spata2_obj_10x_2_log_normalised_sct@data[["GSM5691529_Sample_G21d"]][["scaled"]]), breaks = 100, main = "Distribution of log Normalized Counts with SCT = TRUE")

exp_matrix_log_normalised_sct <- getExpressionMatrix(spata2_obj_10x_2_log_normalised_sct)

# head(exp_matrix_log_normalised_sct)

```


## log normalised and Sctransform is list(variable.features.n set to 20000)

```{r log normalised and Sctransform is variable.features.n set to 20000}

# Does sct affect the log distribution

spata2_obj_10x_2_log_normalised_sct_VARIABLE_FT <-
    initiateSpataObject_10X(
        directory_10X = dataset_10x_dir, # the directory from which to load the data
        sample_name = sample_name_spata2_obj,
        SCTransform = list(variable.features.n = 20000), # can be FALSE or diff value
        NormalizeData = TRUE,
        FindVariableFeatures = TRUE,
        ScaleData = TRUE
    )

spata2_obj_10x_2_log_normalised_sct_VARIABLE_FT <- computeGeneMetaData(spata2_obj_10x_2_log_normalised_sct_VARIABLE_FT) 


hist(as.vector(spata2_obj_10x_2_log_normalised_sct_VARIABLE_FT@data[["GSM5691529_Sample_G21d"]][["scaled"]]), breaks = 100, main = "Distribution of log Normalized Counts WITH SCT = list(variable.features.n = 20000)")

exp_matrix_log_normalised_sct_VARIABLE_FT <- getExpressionMatrix(spata2_obj_10x_2_log_normalised_sct_VARIABLE_FT)

# head(exp_matrix_log_normalised_sct_VARIABLE_FT)

```


## SCT is TRUE ONLY, without specifying anthing else for other parameters;

Previous analysis was done like this; Totally incorrect;


```{r SCTransform TRUE ONLY with no other parameters specified}

spata2_obj_10x_2_SCT_ONLY_TRUE <-
     initiateSpataObject_10X(
     directory_10X = dataset_10x_dir, # the directory from which to load the data
     sample_name = sample_name_spata2_obj,
     SCTransform = TRUE
     )
     

spata2_obj_10x_2_SCT_ONLY_TRUE <- computeGeneMetaData(spata2_obj_10x_2_SCT_ONLY_TRUE) 

# Plotting distribution in R
hist(as.vector(spata2_obj_10x_2_SCT_ONLY_TRUE@data[["GSM5691529_Sample_G21d"]][["scaled"]]), breaks = 100, main = "Distribution of SCTransform() Counts")

exp_matrix_SCT_ONLY_TRUE <- getExpressionMatrix(spata2_obj_10x_2_SCT_ONLY_TRUE)

# head(exp_matrix_SCT_ONLY_TRUE)

```

##  SCT is TRUE and others is FALSE

This is the one I will use in the new scripts; SCT inclusive of all genes and without Standard workflow steps


```{r SCT isTRUE and others is FALSE}
spata2_obj_10x_2_SCT_TRUE_others_False <-
    initiateSpataObject_10X(
        directory_10X = dataset_10x_dir, # the directory from which to load the data
        sample_name = sample_name_spata2_obj,
        SCTransform = list(variable.features.n = 20000), # can be FALSE or diff value
        NormalizeData = FALSE,
        FindVariableFeatures = FALSE,
        ScaleData = FALSE
    )

spata2_obj_10x_2_SCT_TRUE_others_False <- computeGeneMetaData(spata2_obj_10x_2_SCT_TRUE_others_False) 

# Plotting distribution in R
hist(as.vector(spata2_obj_10x_2_SCT_TRUE_others_False@data[["GSM5691529_Sample_G21d"]][["scaled"]]), breaks = 100, main = "Distribution of SCTransform() Counts")

exp_matrix_SCT_TRUE_others_False <- getExpressionMatrix(spata2_obj_10x_2_SCT_TRUE_others_False)

# head(exp_matrix_SCT_TRUE_others_False)

```

## SCT TRUE vut varf.n is set to NULL
```{r}
spata2_obj_10x_2_SCT_var_ft_n_Null_others_False <-
    initiateSpataObject_10X(
        directory_10X = dataset_10x_dir, # the directory from which to load the data
        sample_name = sample_name_spata2_obj,
        SCTransform = list(variable.features.n = NULL), # can be FALSE or diff value
        NormalizeData = FALSE,
        FindVariableFeatures = FALSE,
        ScaleData = FALSE
    )

spata2_obj_10x_2_SCT_var_ft_n_Null_others_False <- computeGeneMetaData(spata2_obj_10x_2_SCT_var_ft_n_Null_others_False) 

# Plotting distribution in R
hist(as.vector(spata2_obj_10x_2_SCT_var_ft_n_Null_others_False@data[["GSM5691529_Sample_G21d"]][["scaled"]]), breaks = 100, main = "Distribution of SCTransform() Counts")

exp_matrix_SCT_var_ft_n_Null_others_False <- getExpressionMatrix(spata2_obj_10x_2_SCT_var_ft_n_Null_others_False)


# head(exp_matrix_SCT_var_ft_n_Null_others_False)

```


